---
title: "R Notebook"
author: "senbaikang"
date: "20.03.21"
output:
  pdf_document: default
  html_notebook: default
---

## Constants
```{r}
PREFIX <- "../crc28/improved"
CELL_ORIGIN <- list(
    c("^TP", "Cancer tissue 1"),
    c("^TD", "Cancer tissue 2"),
    c("^TC", "Cancer tissue 3")
  )
PLOT_BRANCH_LENGTH <- TRUE
PLOT_HPD <- TRUE
SCALE_LONGEST_BRANCH <- 2

SIEVE_VCF <- paste(PREFIX, "geneannotator/sieve_vcf.rds", sep = "/")
MONOVAR <- "../crc28/crc28.monovar.vcf.gz"
MONOVAR_VCF <- paste(PREFIX, "geneannotator/monovar_filtered_vcf.rds", sep = "/")
```

## Load helper functions
```{r warning=FALSE,message=FALSE}
source(file = "real_data_visualization.R")
```

## Visualize tree

### Load local tree and get tips
```{r}
tree <- read.beast(paste(PREFIX, "geneannotator/annotated.tree", sep = "/"))

tip.df <- get.tip.info(tree, pat2names = CELL_ORIGIN)

ordered.tips <- get.ordered.tip.names(tree)
```

### Plot tree with nodes
```{r}
tree.branch.size <- 2.0 / .pt
tree.out.format <- cairo_pdf
tree.out.dpi <- 300
tree.out.limitsize <- FALSE
tip.point.size <- 4 / .pt
tip.label.align <- TRUE
tip.label.linetype <- "dotted"
tip.label.linesize <- 1 / .pt
tip.label.size <- 6 / .pt
tip.label.color <- get.tip.color(1)
tip.label.legend.title <- "Cell origin"
legend.spacing <- unit(0, "pt")
legend.key.size <- unit(3, "mm")
legend.text.size <- element_text(size = 6)
lengend.title.size <- element_text(size = 6)

mark.complex.genes <- list(
  general = TRUE,
  strict = TRUE
)
gene.dup.gene.sep <- " \U00D7 "
gene.sep.criterion <- tibble(
    num = c(2, 20, 40, 60, 80, 100, 120, 140),
    group = c(1, 2, 3, 4, 5, 6, 7, 8)
  )
gene.sep.collapse <- ","
gene.text.alpha <- 0.9
gene.text.size <- 5 / .pt
gene.text.repel.min.segment.length <- 0
gene.text.repel.max.overlaps <- 10
gene.text.repel.label.padding <- unit(0.1, "lines")
gene.text.repel.point.padding <- unit(0, "lines")
gene.label.repel.label.r <- unit(0.1, "lines")
gene.label.repel.label.size <- unit(0.15, "mm")
gene.legend.title <- "Mutation event"
gene.repel.type <- "label" # "label" or "text"
gene.text.color.type <- "color" # "fill" or "color" / "colour"
```

Plot a tree with HPD.
```{r}
if (PLOT_HPD) {
  plot.nodes(
    tree.plot = get.ggtree(
      tree,
      branch.alpha = 1.0,
      plot.branch.length = TRUE,
      branch.size = tree.branch.size,
      scale.longest.branch = 1L
    ),
    anno.tip.text = FALSE,
    anno.tip.label = TRUE,
    tip.data = tip.df,
    tip.point.size = tip.point.size,
    tip.label.align = tip.label.align,
    tip.label.linetype = tip.label.linetype,
    tip.label.linesize = tip.label.linesize,
    tip.label.offset = 0.0000001,
    tip.label.size = tip.label.size,
    tip.label.legend.title = tip.label.legend.title,
    tip.label.color = tip.label.color,
    x.limit = c(0, 0.000018),
    legend.spacing = legend.spacing,
    legend.pos = c(0, 1),
    legend.just = c("left", "top"),
    legend.direction = "vertical",
    legend.key.size = legend.key.size,
    legend.text.size = legend.text.size,
    lengend.title.size = lengend.title.size,
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    anno.hpd = TRUE,
    hpd.color = "red",
    hpd.size = 3 / .pt,
    hpd.alpha = 1.0,
    anno.branch.length = FALSE,
    branch.length.digit = 3L,
    branch.length.text.size = 6L / .pt,
    branch.length.color = "lightblue",
    branch.length.alpha = 1.0,
    label.repel.arrow.length = unit(1.2, "mm"),
    label.repel.force = 1,
    label.repel.force_pull = 1,
    label.repel.min.segment.length = 0,
    label.repel.max.overlaps = 100,
    label.repel.label.padding = unit(0.1, "lines"),
    label.repel.box.padding = unit(0.1, "lines"),
    label.repel.label.r = unit(0.1, "lines"),
    label.repel.label.size = unit(0.15, "mm"),
    anno.posterior = FALSE,
    seed = 1L,
    out.file = paste(PREFIX, "geneannotator/tree_HPD.pdf", sep = "/"),
    out.format = tree.out.format,
    out.width = 183,
    out.height = 100,
    out.dpi = tree.out.dpi,
    out.limitsize = tree.out.limitsize
  )
}
```

Plot a tree with genes annotated.
```{r}
tree.plot <- get.ggtree(
  tree,
  plot.branch.length = PLOT_BRANCH_LENGTH,
  branch.size = tree.branch.size,
  scale.longest.branch = SCALE_LONGEST_BRANCH
)
tree.plot <- plot.nodes(
  tree.plot = tree.plot,
  tip.data = tip.df,
  tip.point.size = tip.point.size,
  tip.label.align = tip.label.align,
  tip.label.linetype = tip.label.linetype,
  tip.label.linesize = tip.label.linesize,
  tip.label.offset = 0.0000001,
  tip.label.size = tip.label.size,
  tip.label.legend.title = tip.label.legend.title,
  tip.label.color = tip.label.color,
  x.limit = c(0, 0.0000055),
  legend.spacing = legend.spacing,
  legend.pos = c(0, 0.25),
  legend.just = c("left", "top"),
  legend.direction = "vertical",
  legend.key.size = legend.key.size,
  legend.text.size = legend.text.size,
  lengend.title.size = lengend.title.size,
  plot.margin = unit(c(0, 0, 0, 0), "mm"),
  posterior.threshold = 0.5,
  internal.node.text.size = 6 / .pt,
  internal.node.label.padding = unit(0.1, "lines"),
  internal.node.label.r = unit(0.1, "lines"),
  internal.node.label.size = unit(0.15, "mm"),
  internal.node.label.alpha = 1.0,
  internal.node.label.fill.color = "white"
)
tree.plot
```

1. All genes.
```{r}
tree.plot.1 <- plot.genes(
  tree.plot,
  mode = 1L,
  mark.complex.genes = mark.complex.genes,
  dup.gene.sep = gene.dup.gene.sep,
  criterion = gene.sep.criterion,
  collapse = gene.sep.collapse,
  text.alpha = gene.text.alpha,
  text.size = gene.text.size,
  # arrow.length = unit(1, "mm"),
  # arrow.type = "open",
  text.repel.force = 1,
  text.repel.force_pull = 1,
  text.repel.min.segment.length = gene.text.repel.min.segment.length,
  text.repel.max.overlaps = 100,
  text.repel.label.padding = gene.text.repel.label.padding,
  text.repel.box.padding = unit(0.4, "lines"),
  text.repel.point.padding = gene.text.repel.point.padding,
  label.repel.label.r = gene.label.repel.label.r,
  label.repel.label.size = gene.label.repel.label.size,
  legend.title = gene.legend.title,
  repel.type = gene.repel.type, # "label" or "text"
  text.color.type = gene.text.color.type, # "fill" or "color" / "colour"
  direction = "y",
  seed = 10L,
  out.file = paste0(PREFIX, "/geneannotator/tree_all.pdf"),
  out.format = tree.out.format,
  out.width = 183,
  out.height = 110,
  out.dpi = tree.out.dpi,
  out.limitsize = tree.out.limitsize
)
tree.plot.1
```

2. ISA genes.
```{r}
# tree.plot.2 <- plot.genes(
#   tree.plot,
#   mode = 2L,
#   dup.gene.sep = gene.dup.gene.sep,
#   criterion = gene.sep.criterion,
#   collapse = gene.sep.collapse,
#   text.alpha = gene.text.alpha,
#   text.size = gene.text.size,
#   # arrow.length = unit(1, "mm"),
#   # arrow.type = "open",
#   text.repel.force = 1,
#   text.repel.force_pull = 1,
#   text.repel.min.segment.length = gene.text.repel.min.segment.length,
#   text.repel.max.overlaps = 100,
#   text.repel.label.padding = gene.text.repel.label.padding,
#   text.repel.box.padding = unit(0.1, "lines"),
#   text.repel.point.padding = gene.text.repel.point.padding,
#   label.repel.label.r = gene.label.repel.label.r,
#   label.repel.label.size = gene.label.repel.label.size,
#   legend.title = gene.legend.title,
#   repel.type = gene.repel.type, # "label" or "text"
#   text.color.type = gene.text.color.type, # "fill" or "color" / "colour"
#   seed = 6000L,
#   out.file = paste0(PREFIX, "/geneannotator/tree_isa.pdf"),
#   out.format = tree.out.format,
#   out.width = 180,
#   out.height = 80,
#   out.dpi = tree.out.dpi,
#   out.limitsize = tree.out.limitsize
# )
# tree.plot.2
```

3. FSA genes.
```{r}
# tree.plot.3 <- plot.genes(
#   tree.plot,
#   mode = 3L,
#   dup.gene.sep = gene.dup.gene.sep,
#   criterion = gene.sep.criterion,
#   collapse = gene.sep.collapse,
#   text.alpha = gene.text.alpha,
#   text.size = gene.text.size,
#   # arrow.length = unit(1, "mm"),
#   # arrow.type = "open",
#   text.repel.force = 1,
#   text.repel.force_pull = 1,
#   text.repel.min.segment.length = gene.text.repel.min.segment.length,
#   text.repel.max.overlaps = 10,
#   text.repel.label.padding = gene.text.repel.label.padding,
#   text.repel.box.padding = unit(0.1, "lines"),
#   text.repel.point.padding = gene.text.repel.point.padding,
#   label.repel.label.r = gene.label.repel.label.r,
#   label.repel.label.size = gene.label.repel.label.size,
#   legend.title = gene.legend.title,
#   repel.type = gene.repel.type, # "label" or "text"
#   text.color.type = gene.text.color.type, # "fill" or "color" / "colour"
#   seed = 10L,
#   out.file = paste0(PREFIX, "/geneannotator/tree_fsa.pdf"),
#   out.format = tree.out.format,
#   out.width = 180,
#   out.height = 70,
#   out.dpi = tree.out.dpi,
#   out.limitsize = tree.out.limitsize
# )
# tree.plot.3
```

## Visualize genotype heatmap

### Load data from VCF files
```{r}
if (is.null(SIEVE_VCF) || !file.exists(SIEVE_VCF)) {
  sieveVCF <- loadVcf(
    file = paste(
      PREFIX,
      "results/stage2/variant_calling_dir/none_univ_ormc-fels_fixed_2_ormc/none_univ_ormc-fels_fixed_2_ormc_median.vcf",
      sep = "/"
    ),
    local.copy = SIEVE_VCF,
    cores = 4L
  )
} else {
  sieveVCF <- readRDS(SIEVE_VCF)
}

if(!is.null(MONOVAR)) {
  if (is.null(MONOVAR_VCF) || !file.exists(MONOVAR_VCF)) {
    monovarVCF <- loadVcf(
      file = MONOVAR,
      vcf.geno = c("GT", "GQ"),
      rename.cells = list(
        c("CRC08-", "")
      ),
      included.cells = colnames(sieveVCF),
      included.sites = rownames(sieveVCF),
      destination.file = paste(
        PREFIX,
        "../monovar_filtered.vcf",
        sep = "/"
      ),
      local.copy = MONOVAR_VCF,
      cores = 8L
    )
  } else
    monovarVCF <- readRDS(MONOVAR_VCF)
}
```

### Color mapping function
```{r}
ht_opt(
  "COLUMN_ANNO_PADDING" = unit(0.5, "mm"),
  "fast_hclust" = TRUE,
  "verbose" = TRUE
)

out.hm.width <- 70 # mm
out.hm.height <- 30 # mm
in2mm <- 25.4

gpar.title <- gpar(fontsize = 7L, fontfamily = "sans")
gpar.title.2 <- gpar(fontsize = 7L, fontfamily = "sans")
gpar.text <- gpar(fontsize = 6L, fontfamily = "sans")

color.hm <- get.hm.color(i = 2)
hm.row.title <- "Cells"
hm.row.title.side <- "left"
hm.row.title.gp <- gpar.title.2
hm.column.title <- "SNV sites"
hm.column.title.side <- "top"
hm.column.title.gp <- gpar.title.2
hm.row.order <- ordered.tips
hm.row.names.side <- "left"
hm.row.names.gp <- gpar.text
```

### Genotype heatmap from Sieve
Cells are ordered according to the above tree, and sites are ordered with a specified distance method.
```{r}
sieve.gt.data <- get.ternary(sieveVCF)
sieve.gt.legend <- get.legend(sieve.gt.data, is.ado = FALSE)

sieve.gt.hm <- Heatmap(
  t(sieve.gt.data),
  col = color.hm,
  row_title = hm.row.title,
  row_title_side = hm.row.title.side,
  row_title_gp = hm.row.title.gp,
  column_title = hm.column.title,
  column_title_side = hm.column.title.side,
  column_title_gp = hm.column.title.gp,
  cluster_rows = FALSE,
  row_order = hm.row.order,
  row_names_side = hm.row.names.side,
  row_names_gp = hm.row.names.gp,
  row_names_centered = TRUE,
  show_column_dend = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  heatmap_legend_param = list(
    at = as.integer(names(sieve.gt.legend)),
    labels = sieve.gt.legend,
    title = "Genotype",
    grid_height = unit(3, "mm"),
    grid_width = unit(3, "mm"),
    color_bar = "discrete",
    border = "black",
    title_gp = gpar.title,
    labels_gp = gpar.text
  )
)

draw(sieve.gt.hm)

sieve.gt.hm.column.order <- column_order(sieve.gt.hm)

gt.key.sites <- columnAnnotation(
    labelled_sites = anno_mark(
      at = c(
        sieve.gt.hm.column.order[1L],
        sieve.gt.hm.column.order[2000L],
        sieve.gt.hm.column.order[4000L],
        sieve.gt.hm.column.order[6000L],
        sieve.gt.hm.column.order[length(sieve.gt.hm.column.order)]
        ),
      labels = get.formatted.num.as.char(c(
        "1",
        "2000",
        "4000",
        "6000",
        as.character(length(sieve.gt.hm.column.order))
        )),
      which = "column",
      side = "top",
      labels_gp = gpar.text,
      labels_rot = 0,
      link_width = unit(1, "pt"),
      link_height = unit(2, "pt")
    ),
    show_legend = FALSE
  )

sieve.gt.hm.column.order <- rownames(sieveVCF)[sieve.gt.hm.column.order]

cairo_pdf(
  paste(PREFIX, "geneannotator/gt_sieve.pdf", sep = "/"),
  width = out.hm.width / in2mm,
  height = out.hm.height / in2mm,
  onefile = FALSE
)

sieve.gt.hm <- Heatmap(
  t(sieve.gt.data),
  col = color.hm,
  row_title = hm.row.title,
  row_title_side = hm.row.title.side,
  row_title_gp = hm.row.title.gp,
  column_title = hm.column.title,
  column_title_side = hm.column.title.side,
  column_title_gp = hm.column.title.gp,
  cluster_rows = FALSE,
  row_order = hm.row.order,
  column_order = sieve.gt.hm.column.order,
  row_names_side = hm.row.names.side,
  row_names_gp = hm.row.names.gp,
  row_names_centered = TRUE,
  show_column_dend = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  heatmap_legend_param = list(
    at = as.integer(names(sieve.gt.legend)),
    labels = sieve.gt.legend,
    title = "Genotype",
    grid_height = unit(3, "mm"),
    grid_width = unit(3, "mm"),
    color_bar = "discrete",
    border = "black",
    title_gp = gpar.title,
    labels_gp = gpar.text
  ),
  top_annotation = gt.key.sites,
  use_raster = FALSE
)
  
draw(sieve.gt.hm)

dev.off()
```

### ADO state heatmap from Sieve
Cells are ordered according to the above tree, and sites are ordered with a specified distance method.
```{r}
# sieve.ado.data <- get.ado.state(sieveVCF)
# ado.legend <- get.legend(sieve.ado.data, is.ado = TRUE)
#
# # sieve.ado.hm <- Heatmap(
# #   t(sieve.ado.data),
# #   col = color.hm,
# #   row_title = hm.row.title,
# #   row_title_side = hm.row.title.side,
# #   row_title_gp = hm.row.title.gp,
# #   column_title = hm.column.title,
# #   column_title_side = hm.column.title.side,
# #   column_title_gp = hm.column.title.gp,
# #   cluster_rows = FALSE,
# #   row_order = hm.row.order,
# #   row_names_side = hm.row.names.side,
# #   row_names_gp = hm.row.names.gp,
# #   row_names_centered = TRUE,
# #   show_column_dend = FALSE,
# #   show_column_names = FALSE,
# #   heatmap_legend_param = list(
# #     at = as.integer(names(ado.legend)),
# #     labels = ado.legend,
# #     title = "ADO state",
# #     grid_height = unit(3, "mm"),
# #     grid_width = unit(3, "mm"),
# #     color_bar = "discrete",
# #     border = "black",
# #     title_gp = gpar.title,
# #     labels_gp = gpar.text
# #   )
# # )
# #
# # draw(sieve.ado.hm)
# #
# # sieve.ado.hm.column.order <- column_order(sieve.ado.hm)
# #
# # ado.key.sites <- columnAnnotation(
# #     labelled_sites = anno_mark(
# #       at = c(
# #         sieve.ado.hm.column.order[1L],
# #         sieve.ado.hm.column.order[3000L],
# #         sieve.ado.hm.column.order[6000L],
# #         sieve.ado.hm.column.order[length(sieve.ado.hm.column.order)]
# #         ),
# #       labels = get.formatted.num.as.char(c(
# #         "1",
# #         "3000",
# #         "6000",
# #         as.character(length(sieve.ado.hm.column.order))
# #         )),
# #       which = "column",
# #       side = "top",
# #       labels_gp = gpar.text,
# #       labels_rot = 0,
# #       link_width = unit(0.7, "mm"),
# #       link_height = unit(0.7, "mm")
# #     ),
# #     show_legend = FALSE
# #   )
# #
# # sieve.ado.hm.column.order <- rownames(sieveVCF)[sieve.ado.hm.column.order]
#
# cairo_pdf(
#   paste(PREFIX, "geneannotator/ado_sieve.pdf", sep = "/"),
#   width = out.hm.width / in2mm,
#   height = out.hm.height / in2mm,
#   onefile = FALSE
# )
#
# sieve.ado.hm <- Heatmap(
#   t(sieve.ado.data),
#   col = color.hm,
#   row_title = hm.row.title,
#   row_title_side = hm.row.title.side,
#   row_title_gp = hm.row.title.gp,
#   column_title = hm.column.title,
#   column_title_side = hm.column.title.side,
#   column_title_gp = hm.column.title.gp,
#   cluster_rows = FALSE,
#   row_order = hm.row.order,
#   column_order = sieve.gt.hm.column.order,
#   row_names_side = hm.row.names.side,
#   row_names_gp = hm.row.names.gp,
#   row_names_centered = TRUE,
#   show_column_dend = FALSE,
#   show_column_names = FALSE,
#   heatmap_legend_param = list(
#     at = as.integer(names(ado.legend)),
#     labels = ado.legend,
#     title = "ADO state",
#     grid_height = unit(3, "mm"),
#     grid_width = unit(3, "mm"),
#     color_bar = "discrete",
#     border = "black",
#     title_gp = gpar.title,
#     labels_gp = gpar.text
#   ),
#   top_annotation = gt.key.sites,
#   use_raster = FALSE
# )
#
# draw(sieve.ado.hm)
#
# dev.off()
```

### Genotype heatmap from MonoVar
Both rows (sites) and columns (cells) are ordered according to the above Sieve heatmap.
```{r}
if(!is.null(MONOVAR)) {
  monovar.hm.row.order <- get.ordered.intersactions(sieve.gt.hm.column.order, rownames(monovarVCF))
  
  monovar.key.sites <- columnAnnotation(
      labelled_sites = anno_mark(
        at = c(
          monovar.hm.row.order$id[1L],
          monovar.hm.row.order$id[2000L],
          monovar.hm.row.order$id[4000L],
          monovar.hm.row.order$id[6000L],
          monovar.hm.row.order$id[dim(monovar.hm.row.order)[1]]
          ),
        labels = get.formatted.num.as.char(c(
          "1",
          "2000",
          "4000",
          "6000",
          as.character(dim(monovar.hm.row.order)[1])
        )),
        which = "column",
        side = "top",
        labels_gp = gpar.text,
        labels_rot = 0,
        link_width = unit(1, "pt"),
        link_height = unit(2, "pt")
      ),
      show_legend = FALSE
    )

  cairo_pdf(
    paste(PREFIX, "geneannotator/gt_monovar_sieve.pdf", sep = "/"),
    width = out.hm.width / in2mm,
    height = out.hm.height / in2mm,
    onefile = FALSE
    )

  monovar.gt.data <- get.ternary(monovarVCF)
  monovar.gt.legend <- get.legend(monovar.gt.data, is.ado = FALSE)

  monovar.hm <- Heatmap(
    t(monovar.gt.data),
    col = color.hm,
    row_title = hm.row.title,
    row_title_side = hm.row.title.side,
    row_title_gp = hm.row.title.gp,
    column_title = hm.column.title,
    column_title_side = hm.column.title.side,
    column_title_gp = hm.column.title.gp,
    cluster_rows = FALSE,
    row_order = hm.row.order,
    column_order = monovar.hm.row.order$pos,
    row_names_side = hm.row.names.side,
    row_names_gp = hm.row.names.gp,
    row_names_centered = TRUE,
    show_column_dend = FALSE,
    show_row_names = FALSE,
    show_column_names = FALSE,
    heatmap_legend_param = list(
      at = as.integer(names(monovar.gt.legend)),
      labels = monovar.gt.legend,
      title = "Genotype",
      grid_height = unit(3, "mm"),
      grid_width = unit(3, "mm"),
      color_bar = "discrete",
      border = "black",
      title_gp = gpar.title,
      labels_gp = gpar.text
      ),
    top_annotation = monovar.key.sites,
    use_raster = FALSE
  )
  
  monovar.hm <- draw(monovar.hm)
  
  dev.off()
}
```

```{r}
library(xtable)
sieve.gt.stats <- get.genotype.stats(sieveVCF, TRUE)
monovar.gt.stats <- get.genotype.stats(monovarVCF, FALSE)
saveRDS(
  matrix(
    (c(sieve.gt.stats, monovar.gt.stats)),
    nrow = 2,
    byrow = TRUE,
    dimnames = list(
      c("SIEVE", "Monovar"),
      names(sieve.gt.stats)
    )
  ),
  paste(PREFIX, "geneannotator/gt_prop.rds", sep = "/")
)
readRDS(paste(PREFIX, "geneannotator/gt_prop.rds", sep = "/"))
print(xtable(format(readRDS(paste(PREFIX, "geneannotator/gt_prop.rds", sep = "/")), digits = 3L, scientific = TRUE), type = "latex"))
```
